---
title: "Figures and Tables"
date: "January, 2018"
output:
  html_notebook:
    toc: yes
  html_document:
    df_print: paged
    toc: yes
    keep_md: yes
---

```{r}
library(ggbio)
library(EnsDb.Hsapiens.v75)
ensdb <- EnsDb.Hsapiens.v75
library(tidyverse)
library(ggbeeswarm)
library(ggthemes)
library(ggrepel)
library(ggsci)
library(cowplot)
load('reanalysis_processed.Rdata')
```
# Figure 1
```{r, fig.width=6, fig.height=4}
# custom fuction to make data avail for ggplot
grabber <- function(cg_id){
  data <- bVals[cg_id,]
  data <- t(data) %>% data.frame() %>% tibble::rownames_to_column('Sample')
  data$Group <- targets$Sample_Group 
  data <- data %>% gather(Probe, Beta, -Group, -Sample)
  data <- data %>% mutate(Group = case_when(Group == 'Case' ~ 'FAS',
                                            TRUE ~ 'Control'))
  return(data)
}

# volcano
volcano <- DMP_e %>% 
  mutate(Significant = case_when(adj.P.Val < 0.05 ~ 'FDR < 0.05',
                                 TRUE ~ 'Not Significant')) %>% 
  ggplot(aes(x = logFC, y = -log10(P.Value))) +
  scale_color_manual(values = c("#BC3C29FF", "grey")) +
  geom_point(aes(colour=Significant)) +
  geom_label_repel(
    data = DMP_e %>% 
      mutate(Label= case_when(Symbol=='' ~ Probe,
                              TRUE ~ paste(Probe, Symbol, sep=' | '))) %>% 
      filter(adj.P.Val < 0.1 & grepl('MAP3K13', Symbol) | adj.P.Val < 0.1 & grepl('GFI1', Symbol)),
    aes(label = Label),
    alpha=0.5,
    force=10,
    max.iter = 100000) +
  coord_cartesian(xlim=c(-0.75,0.75)) +
  ylab('-log10(P value)')

# plot most significantly differentially methylated CpGs
top_CG_up <- DMP_e  %>% filter(adj.P.Val < 0.1 & grepl('MAP3K13', Symbol)) %>% arrange(P.Value) %>%  pull(Probe)
top_CG_down <- DMP_e  %>% filter(adj.P.Val < 0.1 & grepl('GFI1', Symbol)) %>% arrange(P.Value) %>%  pull(Probe)
up <- grabber(top_CG_up)
down <- grabber(top_CG_down)

up_plot <- ggplot(up, aes(x=Group, y=Beta, colour=Group)) + 
  facet_wrap(~Probe, ncol=10, strip.position = 'bottom') + 
  geom_boxplot(width=0.4, outlier.shape = NA) +
  geom_quasirandom(size=0.5, alpha=0.4) +
  theme_minimal() +
  ggsci::scale_colour_nejm() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  theme(strip.text.x = element_text(angle=45)) + 
  coord_cartesian(ylim=c(0,1))

down_plot <- ggplot(down, aes(x=Group, y=Beta, colour=Group)) + 
  facet_wrap(~Probe, ncol=10, strip.position = 'bottom') + 
  geom_boxplot(width=0.4, outlier.shape = NA) +
  geom_quasirandom(size=0.5, alpha=0.4) +
  theme_minimal() +
  ggsci::scale_colour_nejm() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  theme(strip.text.x = element_text(angle=45)) + 
  coord_cartesian(ylim=c(0,1))

legend <- get_legend(up_plot + theme(legend.position="left"))
box_plots <- plot_grid(up_plot + theme(legend.position="none"), 
                       down_plot + theme(legend.position="none"), 
                       nrow=2, align = 'h', labels = c('B','C'))
plot_grid(volcano, box_plots, legend, nrow=1, labels=c('A',NULL,NULL),rel_widths = c(2,1,0.2))

```

# Figure 2
```{r, fig.width=6, fig.height=2}
gfi1 <- DMP_e %>% filter(chr=='chr1', pos > 92940000, pos < 92955000) %>% arrange(chr, pos) %>% pull(Probe)
gfi1_b <- grabber(gfi1)
gfi1_b <- left_join(gfi1_b, DMP_e %>% filter(grepl('GFI1$', Symbol)), by=c('Probe')) 
gfi1_meaned <- gfi1_b %>% group_by(Group, Probe) %>% summarise(Beta = mean(Beta), pos=mean(pos))
gfi1_plot <- gfi1_b %>% 
  ggplot(aes(x=pos, y=Beta, group=Group, colour=Group))  + 
  geom_jitter(width=100, alpha=0.01) + 
  coord_cartesian(ylim=c(0,1)) + 
  geom_line(data = gfi1_meaned %>% data.frame(), aes(x=pos,y=Beta,colour=Group), size=0.5) + 
  scale_color_nejm() +
  coord_cartesian(xlim=c(92940000,92955000)) + xlab('chr1')

map13_start <- 185000000
map13_stop <- 185010000
map13 <- DMP_e %>% filter(chr=='chr3', pos > map13_start, pos < (map13_stop+1000000)) %>% arrange(chr, pos) %>% pull(Probe)
map13 <- grabber(map13)
map13<- left_join(map13, DMP_e %>% filter(chr=='chr3', pos > map13_start, pos < (map13_stop+1000000)), by=c('Probe')) 
map13_meaned <- map13 %>% group_by(Group, Probe) %>% summarise(Beta = mean(Beta), pos=mean(pos))
map13_plot <- map13 %>% 
  ggplot(aes(x=pos, y=Beta, group=Group, colour=Group))  + 
  geom_jitter(width=10, alpha=0.01) + 
  coord_cartesian(ylim=c(0,1)) + 
  geom_line(data = map13_meaned %>% data.frame(), aes(x=pos,y=Beta,colour=Group), size=0.5) + 
  scale_color_nejm() +
  coord_cartesian(xlim=c(map13_start,map13_stop)) + xlab('chr3')


gfi_gene <- ggplot() + geom_alignment(ensdb, which = TxNameFilter("ENST00000427103"), names.expr=NULL, stat='identity') +
  coord_cartesian(xlim=c(92940000,92955000)) + xlab('GFI1')
map3_gene <- ggplot() + geom_alignment(ensdb, which = TxNameFilter("ENST00000424227"), names.expr=NULL, stat='identity') +
  coord_cartesian(xlim=c(map13_start,map13_stop)) + xlab('MAP3K13')


plot_grid(gfi1_plot + theme(legend.position="none"), map13_plot, gfi_gene, map3_gene, labels = c('A','B',NULL,NULL), ncol=2, align='v', axis = 'lr', rel_heights = c(1,0.3))
```

# GO enrichment
Not much for GO enrichment when using Enrichr (only uses direct-ish overlaps between probes and genes)

LOADS of stuff (but anything useful?) when using GREAT (which takes all probes and assigns closest gene)
```{r}
# paste these lists into enrichr (http://amp.pharm.mssm.edu/Enrichr/)
DMP_e %>% filter(adj.P.Val < 0.05, logFC > 0) %>% pull(Symbol) %>% unique() %>% paste(., collapse = ',')

DMP_e %>% filter(adj.P.Val < 0.05, logFC < 0) %>% pull(Symbol) %>% unique() %>% paste(., collapse = ',')

# for GREAT (http://great.stanford.edu)
sig <- DMP_e %>% dplyr::select(Probe, chr, pos, Gene=Symbol, Type, logFC,FDR = adj.P.Val) %>% filter(FDR < 0.05) %>% mutate(Type = gsub('_',' ', Type)) %>% mutate(Type=gsub('hg19 genes ','',Type)) %>% mutate(Type=gsub('hg19 cpg ','',Type)) %>% mutate(pos2=pos+1) %>% dplyr::select(chr, pos, pos2) %>% data.frame()
write_tsv(sig, path='sig.bed', col_names = F)
background <- DMP_e %>% mutate(pos2=pos+1) %>% dplyr::select(chr, pos, pos2)
write_tsv(background %>% arrange(chr, pos) %>% mutate(pos2=as.integer(pos2)),path = 'background.bed', col_names = FALSE)
```


# Table 2
Currently filtering at FDR < 0.1.
In paper with FDR < 0.01 (11 probes)?
```{r}
table2_1 <- DMP_e %>% dplyr::select(Probe, chr, pos, Gene=Symbol, Type, logFC,FDR = adj.P.Val) %>% filter(FDR < 0.1) %>% mutate(Type = gsub('_',' ', Type)) %>% mutate(Type=gsub('hg19 genes ','',Type)) %>% mutate(Type=gsub('hg19 cpg ','',Type))
table2_1
write.csv(table2_1, file='figures/table2.csv')

```

# Kobor validation 
## Table 3
```{r}
kobor_sig <- readxl::read_xlsx('13072_2016_74_MOESM2_ESM.xlsx')
kobor_probes <- kobor_sig %>% pull(Probe)

kobor_sig2 <- readxl::read_xlsx('13148_2018_439_MOESM1_ESM.xlsx', sheet = 'Supplementary table 2')
kobor_probes_validate2 <- kobor_sig2 %>% pull(CpG)
# number of Kobor probes in our dataset
overlap_num <- DMPs %>% tibble::rownames_to_column('Probe') %>% filter(Probe %in% kobor_probes_validate2) %>% nrow()

DMP_e %>% filter(Probe %in% kobor_probes, P.Value < (0.05/overlap_num)) %>% dplyr::select(Probe, chr, pos, Gene=Symbol, Type, logFC,FDR = adj.P.Val)  %>% mutate(Type = gsub('_',' ', Type)) %>% mutate(Type=gsub('hg19 genes ','',Type)) %>% mutate(Type=gsub('hg19 cpg ','',Type))
```

## Supplementary Figure 1
```{r}


z <- DMPs %>% 
  tibble::rownames_to_column('Probe') %>% 
  mutate(Kobor_sig = case_when(Probe %in% kobor_probes_validate2 ~ 'Validated',
                               TRUE ~ 'No')) 


t.test(z$logFC ~ z$Kobor_sig)

z <- DMPs %>% 
  tibble::rownames_to_column('Probe') %>% 
  mutate(Kobor_sig = case_when(Probe %in% kobor_probes ~ 'Validated',
                               TRUE ~ 'No')) 


t.test(z$logFC ~ z$Kobor_sig)

DMP_e %>% 
 mutate(`Kobor Probes` = case_when(Probe %in% kobor_probes_validate2 ~ 'Validated',
                               Probe %in% kobor_probes ~ 'Significant',
                               TRUE ~ 'Not Significant')) %>% 
  ggplot(aes(x=abs(logFC), colour=`Kobor Probes`)) + 
  geom_density() +
  theme_bw() +
  scale_color_nejm() + coord_cartesian(xlim=c(0,0.5))
```
