---
title: "Prep beta values for Kobor model "
output: html_notebook
---
```{r}
# package loading
library(data.table)
library(tidyverse)
library(annotatr)
library(ggthemes)
library(ggrepel)
library(ggsci)
library(cowplot)
library(limma)
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(IlluminaHumanMethylation450kmanifest)
library(GenomicFeatures)
library(RColorBrewer)
library(missMethyl)
library(matrixStats)
library(Gviz)
library(DMRcate)
library(stringr)
library(data.table)


sessionInfo()
```

# What this does
Create csv file(s) for https://fasdpredictor.shinyapps.io/fasdpredictorapp/

## Pull in annotation, metadata, and raw Illumina 450k data from the idat files

```{r}
# pull annotation
ann450k <-  getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

# bring in raw data (Illumina .idat)
# data and code available at:
#   https://github.com/davemcg/fetal_alcohol
#   GSE65428
metadata <- fread('~/git/fetal_alcohol/sentrix_id_conversion_table_demographics_v6.txt')
targets <- read.metharray.sheet("/Volumes/ThunderBay/PROJECTS/brody/fetal_alcohol/Tsai_Sample_083/",pattern="Sheet.csv$")
#targets <- read.metharray.sheet("~/Desktop/fetal_alcohol/Tsai_Sample_083/",pattern="Sheet.csv$")
rgSet <- read.metharray.exp(targets=targets)
sampleNames(rgSet) <- rgSet$Sample_Name

# add sample group info to targets, fewer rows in metadata, already parsed down to remove outliers
targets <- left_join(data.frame(metadata),targets,by=c("Sample"="Sample_Name"))
targets$Sample_Group <- targets$Case.Control
targets$Sample_Name <- targets$Sample

# remove samples based on previous outlier analysis / ethnicity (see Methods of "DNA methylation analyses in archived newborn bloodspots reveal signals associated with fetal alcohol syndrome, maternal smoking, gender, and ethnicity")
keep <-  rgSet$Sample_Name %in% targets$Sample_Name
rgSet <- rgSet[,keep]


```


## Normalize data with preprocessQuantile 

```{r, fig.width=2}
# normalize the data; this results in a GenomicRatioSet object
mSetSq <- preprocessQuantile(rgSet)

```


## Calculate M-values for statistical analysis and Beta values for plotting

```{r}
bVals <- getBeta(mSetSq)
#head(bVals[,1:5])
```

# explicitly match targets sample order (rows) with mVars
```{r, fig.height=2, fig.width=4}
targets <- targets[match(colnames(bVals), targets$Sample),]
```


# Create bVals csv in kobor specific order, as downloaded from their shiny app
```{r}
probes <- fread('data/cpgs2input_fasdpredictor.txt')
bVals_kobor_order <- bVals[probes$x,]
write_csv(bVals_kobor_order %>% t() %>% data.frame() %>% rownames_to_column(''), path='bVals_kobor_probes.csv')
```

# Write metadata (which are FAS / control)
```{r}
write_csv(targets %>% dplyr::select(Sample_Group) %>% mutate(Sample_Group = case_when(Sample_Group=='Case'  ~ 'FASD', TRUE ~ Sample_Group)), path='bVals_kobor_probes_metadata.csv')
```